---
title: "VTPEH6270-CP04"
author: "Andrew Rodman"
date: "2026-03-06"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Introduction

## 1.1 Packages 

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)

```

## 1.2 Data Cleaning, Merging and Deriving

```{r, warning=FALSE}
setwd("C:/Users/andre/OneDrive/Documents/VTPEH6270/")
fludata = read.csv("Data/Influenza_Laboratory-Confirmed_Cases_by_County__Beginning_2009-10_Season_20260202.csv")
svi = read.csv("Data/SVI_2016_US_county.csv")

flu_svi = merge(fludata, svi, by= "FIPS", all.x = T)
flu_svi$Count <- as.numeric(flu_svi$Count)
flu_svi$Week.Ending.Date <- as.Date(flu_svi$Week.Ending.Date, format = "%m/%d/%Y")
flu_svi$Season_Year <- as.numeric(substr(flu_svi$Season, 6, 9))
flu_svi = flu_svi[, c("Season_Year",
                      "Region",
                      "County",
                      "Week.Ending.Date",
                      "Disease",
                      "Count",
                      "E_TOTPOP",
                      "EP_POV",
                      "E_PCI",
                      "EP_UNEMP",
                      "FIPS")]
flu_svi <- flu_svi %>%
  rename(Total_Pop = E_TOTPOP,
         Poverty_Rate = EP_POV,
         Income_Per_Capita = E_PCI,
         Date = Week.Ending.Date,
         Unemployment_Rate = EP_UNEMP)

flu_svi_year <- flu_svi %>%
group_by(Season_Year, County) %>%
summarise(
Region = first(Region),
Total_Count = sum(Count),
Total_Pop = first(Total_Pop),
Poverty_Rate = first(Poverty_Rate),
Income_Per_Capita = first(Income_Per_Capita),
Unemployment_Rate = first(Unemployment_Rate),
FIPS = first(FIPS),
.groups = "drop"
)
flu_svi_year <- flu_svi_year %>%
mutate(Incidence_Rate = (Total_Count / Total_Pop) * 100000)

```

# 2. Objective and Approach

## 2.1 Specific Objective

Do annual influenza incidence rates differ in New York State counties by income?

## 2.2 Data Description

```{r}

variable_table <- data.frame(
  Name = c("County", "Income_Per_Capita", "Incidence_Rate", "Season_Year"),
  Type = c("Categorical", "Continuous", "Continuous", "Discrete"),
  Class = c("Character", "Numeric", "Numeric", "Date")
)

knitr::kable(variable_table, caption = "Variable Types and Classes")

```

## 2.3 Data Visualization

```{r, warning=FALSE, message=FALSE}
ggplot(flu_svi_year, aes(x = Income_Per_Capita, y = Incidence_Rate)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Association Between Income Per Capita and Incidence Rate",
    x = "Income Per Capita",
    y = "Incidence Rate"
  ) +
  theme_minimal()
```

## 2.4 Description of Plausible Relationships

Incidence Rate and Income Per Capita seem to have a slightly negative linear
relationship.

## 2.5 Parameters of Interest 

```{r}
parameters <- data.frame(
  Parameter = c("Intercept", "Slope (Effect Size)", "Error / Noise"),
  Symbol = c("a", "b", "epsilon"),
  Description = c(
    "Expected Incidence Rate when Income Per Capita = 0",
    "Change in Incidence Rate for a one-unit increase in Income Per Capita",
    "Random variation in Incidence Rate not explained by Income Per Capita"))

knitr::kable(parameters, caption = "Parameters")

model <- lm(Incidence_Rate ~ Income_Per_Capita, data = flu_svi_year)
mean_income <- mean(flu_svi_year$Income_Per_Capita, na.rm = TRUE)
sd_income <- sd(flu_svi_year$Income_Per_Capita, na.rm = TRUE)
```
a = 956
b = -0.011
mean_income = 28987.5
sd_income = 7366

# 3. Simulation

## 3.1 Simulation Basis 

```{r}

set.seed(123)
n <- 200
Income_Sim <- rnorm(n, mean = 28987.5, sd = 7366)
a <- 956 
b <- -0.011
epsilon <- rnorm(n, mean = 0, sd = 50)
Incidence_Sim <- a + b * Income_Sim + epsilon
sim_data <- data.frame(
  Income_Sim = Income_Sim,
  Incidence_Sim = Incidence_Sim
)

```
## 3.2 Simulation Output

```{r}
sim_data$Income_Quartile <- cut(
  sim_data$Income_Sim,
  breaks = quantile(sim_data$Income_Sim, probs = seq(0, 1, 0.25)),
  include.lowest = TRUE,
  labels = c("Q1", "Q2", "Q3", "Q4")
)
summary_stats <- sim_data %>%
  group_by(Income_Quartile) %>%
  summarise(
    n = n(),
    Mean = mean(Incidence_Sim),
    Median = median(Incidence_Sim),
    SD = sd(Incidence_Sim),
    Min = min(Incidence_Sim),
    Max = max(Incidence_Sim)
  )

knitr::kable(summary_stats, caption = "Quartile Summary Statistics")
```

## 3.3 Simulation Function
```{r}
run_simulation <- function(
  effect_size = -0.011,
  noise       = 1.0,
  n           = 1000,
  a           = 956,
  mean_income = 28987.5,
  sd_income   = 7366,
  seed        = 42
) {
  set.seed(seed)
  sim_data <- data.frame(
    Income_Sim = rnorm(n, mean = mean_income, sd = sd_income)
  )
  sim_data$Incidence_Sim <- a + effect_size * sim_data$Income_Sim + rnorm(n, mean = 0, sd = noise)
  sim_data$Income_Quartile <- cut(
    sim_data$Income_Sim,
    breaks = quantile(sim_data$Income_Sim, probs = seq(0, 1, 0.25)),
    include.lowest = TRUE,
    labels = c("Q1", "Q2", "Q3", "Q4")
  )
  summary_stats <- sim_data %>%
    group_by(Income_Quartile) %>%
    summarise(
      n                = n(),
      Mean_Incidence   = mean(Incidence_Sim),
      Median_Incidence = median(Incidence_Sim),
      SD_Incidence     = sd(Incidence_Sim),
      Min_Incidence    = min(Incidence_Sim),
      Max_Incidence    = max(Incidence_Sim),
      .groups = "drop"
    )
  list(
    params        = list(effect_size = effect_size,
                         noise       = noise,
                         n           = n),
    sim_data      = sim_data,
    summary_stats = summary_stats,
    table         = knitr::kable(summary_stats, caption = "Quartile Summary Statistics")
  )
}

```
## 3.4 Simulation Automation

```{r}
effect_sizes <- c(-0.001, -0.003, -0.005, -0.008, -0.011,
                  -0.015, -0.020, -0.030, -0.050, -0.080)

sample_sizes <- c(50, 100, 250, 500, 750, 1000, 2500, 5000, 10000, 50000)

noise_levels <- c(10, 50, 150)

param_grid <- expand.grid(
  effect_size = effect_sizes,
  n           = sample_sizes,
  noise       = noise_levels
)


simulation_results <- vector("list", nrow(param_grid))

for (i in seq_len(nrow(param_grid))) {
  simulation_results[[i]] <- run_simulation(
    effect_size = param_grid$effect_size[i],
    noise       = param_grid$noise[i],
    n           = param_grid$n[i],
    seed        = i
  )
}

names(simulation_results) <- paste0(
  "es=", param_grid$effect_size,
  "_n=", param_grid$n,
  "_noise=", param_grid$noise
)

all_summaries <- bind_rows(
  lapply(simulation_results, function(x) {
    x$summary_stats %>%
      rename(n_quartile = n) %>%
      mutate(
        effect_size = x$params$effect_size,
        noise       = x$params$noise,
        n           = x$params$sample_size
      )
  })
)

cat("Total simulation runs:",nrow(param_grid),"\n")
```
# 4. Visualization

```{r}
slope_summary <- bind_rows(
  lapply(simulation_results, function(x) {
    if (is.null(x) || is.null(x$sim_data)) return(NULL)
    
    fit <- lm(Incidence_Sim ~ Income_Sim, data = x$sim_data)
    data.frame(
      effect_size = x$params$effect_size,
      sample_size = x$params$n,
      noise       = paste0("Noise = ", x$params$noise),
      est_slope   = coef(fit)[["Income_Sim"]]
    )
  })
)

slope_summary$noise <- factor(slope_summary$noise,
  levels = c("Noise = 10", "Noise = 50", "Noise = 150"))

ggplot(slope_summary,
       aes(x = sample_size, y = est_slope, colour = est_slope)) +
  geom_point(size = 2, alpha = 0.85) +
  geom_line(alpha = 0.4) +
  geom_hline(yintercept = as.numeric(as.character(
    unique(slope_summary$effect_size))),
    linetype = "dashed", colour = "grey60", linewidth = 0.3) +
  facet_grid(noise ~ effect_size) +
  scale_x_log10(
    breaks = c(50, 500, 5000, 50000),
    labels = c("50", "500", "5k", "50k")
  ) +
  scale_colour_viridis_c(option = "plasma", direction = -1,
                         name = "Estimated\nSlope") +
  labs(
    title    = "Estimated Slope by Sample Size, Effect Size and Noise",
    subtitle = "Dashed line = true effect size · x-axis log scaled · rows = noise level · columns = effect size",
    x        = "Sample Size (n, log scale)",
    y        = "Estimated Slope"
  ) +
  theme_minimal(base_size = 9) +
  theme(
    strip.text       = element_text(face = "bold", size = 7),
    plot.title       = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 7),
    legend.position  = "right"
  )
```

# 5. Interpretation

These simulations showcase that the effect is consistently recovered across parameters,
meaning that the estimated slope is always negative, consistent with our original data.
The direction never reverses. Noise also has little influence, as all 3 noise distributions, 
look close. The quartile gradient is also real and detectable, supporting the association between income, 
and influenza incidence. 

# 6. AI Use Disclosure Statement

As part of this assignment, please indicate whether you used any AI-based tools (e.g., ChatGPT, Claude,
Copilot, Gemini, etc.). Indicate:
• Did you use AI? Yes


• If yes: Write a short disclosure statement (2–3 sentences) describing:
• Which tool(s) you used.
Claude and ChatGPT

• How you used it (e.g., to help decide on an analysis approach, to generate a first draft of code, to
improve or debug code you had written yourself, etc.).
Example: “This document was generated using Claude to generate original code, which was then reviewed
and adjusted” or “This document was generate using ChatGPT to assist with debugging of code generated
by the author”.
Please keep your statement brief and honest. The goal is transparency, not detail: we do not need exact
prompts or transcripts, just a clear sense of how AI supported this work

This document was generated using Claude and ChatGPT to formulate a simulation, and produce
graphs and the multi-panel grid. 

